---
output: github_document
bibliography: README.bib
nocite: '@*'
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 8,
  dpi = 120
)
```
# testOTM

<!-- badges: start -->
<!-- badges: end -->

This repository contains a R package which implements the multivariate ranks and quantiles, together with their application in statistical testing, that based on the semi-discrete optimal transportation. The underlying geometric computation is performed by the [Geogram](http://alice.loria.fr/index.php/software/4-library/75-geogram.html) library.

## Installation

<!--
You can install the released version of testOTM from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("testOTM")
```
-->

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
# devtools::install_github("Francis-Hsu/testOTM")
```
## Example

This is a basic example which shows you how to use `testOTM`:

```{r example}
library(testOTM)

# generate some data
n = 100
X = cbind(rnorm(n, 0, 1), rnorm(n, 0, 0.5))

# scale the data to [0, 1] range
X = scaling.min.max(X)

# compute the optimal transport map from U[0, 1]^2 to the data
X.OTM = otm.fit(X)

# plot the restricted Voronoi diagram and the restricted Delaunay triangulation
par(mfrow = c(1, 2))
plot(X.OTM, which = "Both", draw.center = F, draw.map = T)
```

## Reference