---
output: github_document
bibliography: README.bib
nocite: '@*'
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 8,
  fig.height = 4.5,
  dpi = 240
)
```
# testOTM

<!-- badges: start -->
<!-- badges: end -->

This repository contains a R package which implements the multivariate ranks and quantiles, together with their application in statistical testing, that based on the semi-discrete optimal transportation.

## Installation

<!--
You can install the released version of testOTM from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("testOTM")
```
-->
You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
# devtools::install_github("Francis-Hsu/testOTM")
```
## Example

This is a basic example which shows you how to use `testOTM` to visualize the optimal transport map from $U[0, 1]^2$ to a dataset:

```{r example}
library(testOTM)

# generate some data
p = 2
n = 100
Sigma = matrix(c(1, 0, 0, 1), 2, 2)
eS = eigen(Sigma, symmetric = TRUE)
X = t(eS$vectors %*% diag(sqrt(pmax(eS$values, 0)), p) %*% matrix(rnorm(p * n), p))

# scale the data to [0, 1] range
X = scaling.min.max(X)

# compute the optimal transport map from U[0, 1]^2 to the data
X.OTM = tos.fit(X)

# plot the restricted Voronoi diagram and the restricted Delaunay triangulation
par(mfrow = c(1, 2))
plot(X.OTM, which = "Both", draw.center = F, draw.map = T)
```

## Acknowledgements
The author is extremely grateful to Prof. [Bodhisattva Sen](http://www.stat.columbia.edu/~bodhi/Bodhi/Welcome.html) and his student Promit Ghosal for their guidance in the development of this package. The author would also like to thank Dr. [Bruno Lévy](https://members.loria.fr/BLevy/) for his assistance with the [Geogram](http://alice.loria.fr/index.php/software/4-library/75-geogram.html) library, and [TraME](http://www.trame-project.com/) team, whose [`Rgeogram`](https://github.com/TraME-Project/Rgeogram) package provides inspirations in the early build of this package.

## Reference